Natural Selection Across Three Generations of Americans

### 2024-02-28 hrs-selection.qmd
Parents education and improved discussion of generations

# only selected columns to avoid creating suffixes when we join with
# rand.
# Age of oldest kid 2016; age of oldest kid 2010; grandchildren; num siblings who died
rand_fam <- rand_fam |> select(hhid, pn, h13ageokid, h10ageokid, h13gkid, r10sbdied)

@@

rand <- rand |> 
  select(ethnicity, all_of(pgs), all_of(old_pgs), 
         all_of(pcs), weight, raevbrn, rlrs,
         raedyrs, r10sibs, edu, income_resid, income_med, married, 
         agefb, agefbn, rabyear, born, raehsamp, raestrat, 
         parent_weight, ragender, adhd_score, neurot, r10diabe, 
         r10hearte, r10cesd, age_smoked, anxiety, r10cogtot, 
         rlrs_div_h13gkid, child_weight, h13ageokid)

@@

tidy_pgs <- mod_pgs <- list()
tidy_pgs_ed <- mod_pgs_edyrs <- list()
tidy_pgs_sibs <- mod_pgs_sibs <- list()
tidy_pgs_gkids <- mod_pgs_gkids <- list()
tidy_pgs_x_ed <- mod_pgs_x_ed <- list()
tidy_pgs_x_inc <- mod_pgs_x_inc <- list()
tidy_pgs_x_mar <- mod_pgs_x_mar <- list()
tidy_pgs_x_agefb <- mod_pgs_x_agefb <- list()

coefs_fert_edu <- cors_fert_edu <- tidy_pgs_joined <- list()
cors_sibs_edu <- list()
cors_gkids_edu <- list()
tidy_fert_edu_phen <- list()

@@

@@

@@

# The large n_reps is to get an accurate upper tail on the black sample, 
# since it is close to 0:
n_reps <-  599

@@

@@

  cors_fert_edu[[eth]] <- 
    withReplicates(rand_boot, theta = calc_cors_edu, var = "rlrs")
  cors_sibs_edu[[eth]] <- 
    withReplicates(r_parents_boot, theta = calc_cors_edu, var = "r10sibs")
  cors_gkids_edu[[eth]] <- 
    withReplicates(r_child_boot, theta = calc_cors_edu, var = "rlrs_div_h13gkid")
}

@@

ci_boot_fert <- map(cors_fert_edu, \(x) confint(x)[1,])
ci_boot_sibs <- map(cors_sibs_edu, \(x) confint(x)[1,])
ci_boot_gkids <- map(cors_gkids_edu, \(x) confint(x)[1,])

ci_boot_fert_lm <- map(cors_fert_edu, \(x) confint(x)[2,])
ci_boot_sibs_lm <- map(cors_sibs_edu, \(x) confint(x)[2,])
ci_boot_gkids_lm <- map(cors_gkids_edu, \(x) confint(x)[2,])

if (insignificant(ci_boot_fert$white)) warning("ci_boot_fert$white insignificant")
if (significant(ci_boot_fert$black)) warning("ci_boot_fert$black significant")
if (insignificant(ci_boot_sibs$white)) warning("ci_boot_sibs$white insignificant")
if (significant(ci_boot_sibs$black)) warning("ci_boot_sibs$black significant")
if (insignificant(ci_boot_gkids$white)) warning("ci_boot_gkids$white insignificant")
if (significant(ci_boot_gkids$black)) warning("ci_boot_gkids$black significant")
```

@@

We can also examine natural selection in the preceding and succeeding generations, by using reported number of siblings and grandchildren respectively. We regress PGS on respondents' number of living siblings in 2010. Data for dead siblings has too many missing values to use. We reweight respondents by the reciprocal of their number of siblings, to account for parents of many siblings being more likely to be a parent of a respondent. Parents of no siblings cannot be included, so coefficient sizes are not comparable across the generations. Appendix @fig-siblings plots coefficients on number of siblings versus coefficients on years of education. Correlations are significant and negative (correlation `r cors_sibs_edu$white[1]`, bootstrapped 95% C.I. `r ci_boot_sibs$white[1]` to `r ci_boot_sibs$white[2]`). There is a positive and significant correlation across generations, i.e. between PGS coefficients on number of siblings and number of children (`r gen_cor`, bootstrapped 95% C.I. `r ci_gen_cor[1,1]` to `r ci_gen_cor[1,2]`).

@@

Across three generations of Americans, we found estimates of selection to correlate over time and that selection is consistently more positive for traits that negatively predict educational attainment. Patterns of selection appear consistent across the twentieth and early twenty-first century. The effects of natural selection are accumulating and the underlying causes of selection may have stayed the same over time.

>>>>

# only selected columns to avoid creating suffixes when we join with
# rand.
# Age of oldest kid 2016; age of oldest kid 2010; grandchildren; num siblings who died;
# Mother's years of education; father's years of education
rand_fam <- rand_fam |> select(hhid, pn, h13ageokid, h10ageokid, h13gkid, r10sbdied,
                               rameduc, rafeduc) |>
                        mutate(paedyrs = as.numeric(rameduc + rafeduc)/2) #parent ave education

@@

rand <- rand |> 
  select(ethnicity, all_of(pgs), all_of(old_pgs), 
         all_of(pcs), weight, raevbrn, rlrs,
         raedyrs, r10sibs, edu, income_resid, income_med, married, 
         agefb, agefbn, rabyear, born, raehsamp, raestrat, 
         parent_weight, ragender, adhd_score, neurot, r10diabe, 
         r10hearte, r10cesd, age_smoked, anxiety, r10cogtot, 
         rlrs_div_h13gkid, child_weight, h13ageokid, paedyrs)

@@

tidy_pgs <- mod_pgs <- list()
tidy_pgs_ed <- mod_pgs_edyrs <- list()
tidy_pgs_ed_parent <- mod_pgs_edyrs_parent <- list()
tidy_pgs_sibs <- mod_pgs_sibs <- list()
tidy_pgs_gkids <- mod_pgs_gkids <- list()
tidy_pgs_x_ed <- mod_pgs_x_ed <- list()
tidy_pgs_x_inc <- mod_pgs_x_inc <- list()
tidy_pgs_x_mar <- mod_pgs_x_mar <- list()
tidy_pgs_x_agefb <- mod_pgs_x_agefb <- list()

coefs_fert_edu <- cors_fert_edu <- tidy_pgs_joined <- list()
cors_sibs_edu <- list()
cors_sibs_edu_parent <- list()
cors_gkids_edu <- list()
tidy_fert_edu_phen <- list()

@@

  form_pgs_edyrs_parent <- map(pgs, \(x) reformulate(c(x),
                                              response = "paedyrs"))
  mod_pgs_edyrs_parent[[eth]] <- map(form_pgs_edyrs_parent, 
                                     \(x) svyglm_quiet(x, r_parents_eth))
  names(mod_pgs_edyrs_parent[[eth]]) <- pgs

  tidy_pgs_ed_parent[[eth]] <- mod_pgs_edyrs_parent[[eth]] |>  
    map(broom::tidy) |>
    list_rbind() |> 
    filter(term %in% pgs)

@@

  cors_sibs_edu_parent[[eth]] <- cor(tidy_pgs_sibs[[eth]]$estimate, 
                     tidy_pgs_ed_parent[[eth]]$estimate)

@@

# The large n_reps is to get an accurate upper tail on the black sample, 
# since it is close to 0:
n_reps <-  30 #599

@@

calc_cors_edu_parent <- function (weights, data, var) {
  coef_pgs_fert <- pgs |> 
        map(\(x) reformulate(c(x), response = var)) |> 
        map(\(x) lm(x, data, weights = weights)) |> 
        map_dbl(\(x) coef(x)[[2]])

  coef_pgs_edu <- pgs |> 
        map(\(x) reformulate(c(x), response = "paedyrs")) |> 
        map(\(x) lm(x, data, weights = weights)) |> 
        map_dbl(\(x) coef(x)[[2]])

@@

  cors_fert_edu[[eth]] <- 
    withReplicates(rand_boot, theta = calc_cors_edu, var = "rlrs")
  cors_sibs_edu[[eth]] <- 
    withReplicates(r_parents_boot, theta = calc_cors_edu, var = "r10sibs")
  cors_sibs_edu_parent[[eth]] <- 
    withReplicates(r_parents_boot, theta = calc_cors_edu_parent, var = "r10sibs")
  cors_gkids_edu[[eth]] <- 
    withReplicates(r_child_boot, theta = calc_cors_edu, var = "rlrs_div_h13gkid")
}

@@

ci_boot_fert <- map(cors_fert_edu, \(x) confint(x)[1,])
ci_boot_sibs <- map(cors_sibs_edu, \(x) confint(x)[1,])
ci_boot_sibs_parent <- map(cors_sibs_edu_parent, \(x) confint(x)[1,])
ci_boot_gkids <- map(cors_gkids_edu, \(x) confint(x)[1,])

ci_boot_fert_lm <- map(cors_fert_edu, \(x) confint(x)[2,])
ci_boot_sibs_lm <- map(cors_sibs_edu, \(x) confint(x)[2,])
ci_boot_sibs_lm_parent <- map(cors_sibs_edu_parent, \(x) confint(x)[2,])
ci_boot_gkids_lm <- map(cors_gkids_edu, \(x) confint(x)[2,])

if (insignificant(ci_boot_fert$white)) warning("ci_boot_fert$white insignificant")
if (significant(ci_boot_fert$black)) warning("ci_boot_fert$black significant")
if (insignificant(ci_boot_sibs$white)) warning("ci_boot_sibs$white insignificant")
if (insignificant(ci_boot_sibs_parent$white)) warning("ci_boot_sibs_parent$white insignificant")
if (insignificant(ci_boot_sibs_parent$white)) warning("ci_boot_sibs_parent$black significant")
if (significant(ci_boot_sibs$black)) warning("ci_boot_sibs$black significant")
if (insignificant(ci_boot_gkids$white)) warning("ci_boot_gkids$white insignificant")
if (significant(ci_boot_gkids$black)) warning("ci_boot_gkids$black significant")
```

@@

We can also examine natural selection in the preceding and succeeding generations, by using reported number of siblings and grandchildren respectively. We regress PGS on respondents' number of living siblings in 2010. Data for dead siblings has too many missing values to use. We reweight respondents by the reciprocal of their number of siblings, to account for parents of many siblings being more likely to be a parent of a respondent. Parents of no siblings cannot be included, so coefficient sizes are not perfectly comparable across the generations. Appendix @fig-siblings plots coefficients on number of siblings versus coefficients on years of education. Correlations are significant and negative (correlation `r cors_sibs_edu$white[1]`, bootstrapped 95% C.I. `r ci_boot_sibs$white[1]` to `r ci_boot_sibs$white[2]`). There is a positive and significant correlation across generations, i.e. between PGS coefficients on number of siblings and number of children (`r gen_cor`, bootstrapped 95% C.I. `r ci_gen_cor[1,1]` to `r ci_gen_cor[1,2]`).

The HRS has data on the years of education for the parents of respondents. We recalculate the coefficients on education using the parent's PGS, in case the genetics of education and human capital changed between generations. We then correlate these coefficients with the number of siblings, finding a significant result (correlation `r cors_sibs_edu_parent$white[1]`, bootstrapped 95% C.I. `r ci_boot_sibs_parent$white[1]` to `r ci_boot_sibs_parent$white[2]`). The correlation is only slightly smaller and the confidence intervals are only slightly wider. This is surprising since parent's education is likely recorded with error and the respondent's polygenic score is only a proxy for their parents.

@@

Across three generations of Americans, we found estimates of selection to correlate over time and that selection is consistently more positive for traits that negatively predict educational attainment. Patterns of selection appear consistent across the twentieth and early twenty-first century. The result is consistent with that of @Zietsch2014perfect, who found a perfect genetic correlation between number of children and number of grandchildren in a large sample of Swedish twins born between 1915-1929. The underlying causes of selection seem to have stayed the same over time, such that the effects of selection have accumulated.
